Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-23 analysis/paper/paper.Rmd
some very minor typos

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. The native Northern Pikeminnow are one of these predators and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough during the peak emigrations of juvenile Chinook Salmonto be 27,874 in the fall and 19,499 in the spring. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon for varying proportions of fish in their diet, ranging from 30% to 90%. Assuming a median scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, the potential impact of predation on juvenile Chinook Salmon under the median scenario is a reduction in adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Recent studies have examined the downstream movement, distribution, and apparent survival (hereafter 'survival') of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had an approximate 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during fall and early winter months [@Ackerman2018; @Porter2019]. Low survival was attributed to low-velocity water delaying movement rates, thereby increasing predation risk.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area of Deadwater Slough. Length and area were measured in GIS software and using drone-generated orthoimagery.

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. The native Northern Pikeminnow are one of these predators and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough during the peak emigrations of juvenile Chinook Salmon to be 27,874 in the fall and 19,499 in the spring. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon for varying proportions of fish in their diet, ranging from 30% to 90%. Assuming a median scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, the potential impact of predation on juvenile Chinook Salmon under the median scenario is a reduction in adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Recent studies have examined the downstream movement, distribution, and apparent survival (hereafter 'survival') of juvenile salmonid emigrants through the Salmon River, including the Deadwater Slough reach. Sockeye Salmon migrating through the mainstem Salmon River during spring were estimated to have substantially reduced survival in the Deadwater Slough reach [@Axel2015]. Similarly, emigrating DSR Chinook Salmon had an approximate 10% reduction in transition probability through the Deadwater Slough compared to surrounding reaches during fall and early winter months [@Ackerman2018; @Porter2019]. Low survival was attributed to lack of fish cover and low-velocity water delaying movement rates, thereby increasing predation risk.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area of Deadwater Slough. Length and area were measured in QGIS software [@QGISDevelopmentTeam2022] and using drone-generated orthoimagery.
